Mapping Cancer: the Potential of Cartograms and Alternative Map Displays

### 2020-05-22 letter2/submission2.Rmd
Figure numbering and section numbering, and responses to comments on doc

@@

The following reviewer suggestions relate to the introduction:

@@

**Reviewer suggestion:**
**Suggest also noting development of statistical methods and software**
The paragraph that began the *Cancer Atlases* section has been moved to the introduction. This paragraph describes cancer atlases, and the developments oof statistical methods and software.

**Reviewer suggestion:**
**Give the actual reference for Cruikshank 1947**

This quote has been removed.

@@

**Reviewer suggestion:**
**Are you referring to area-based characteristics?**
This socioeconomic indicators mentioned refer to the characteristics of small areas presented in cancer atlases.
The Australian cancer atlas allows filtering using socioeconomic indicators, but this statement has been removed as it is not common in other atlases.

@@

**Reviewer suggestion:**
**Ensure all these refs are included in the list.**
References have been added to credit the original publications of the alternative displays dicussed by Tobler.

>>>>

The first suggestion pertains to the structure of the article. The rest of the suggestions are listed in the order of the comments made in the document  *ACE-19-31-R1-from-the-reviewers*.

@@

**Reviewer suggestion:**
**I would suggest deleting this sentence and the previous one. **
These sentences have been removed

@@

**Reviewer suggestion:**
**Could move to the introduction** and
**Suggest also noting development of statistical methods and software**
The paragraph that began the *Cancer Atlases* section has been moved to the introduction. This paragraph describes cancer atlases, and the developments of statistical methods and software.

**Reviewer suggestion:**
**Give the actual reference for Cruikshank 1947**
This quote has been removed.

@@

**Reviewer suggestion:**
**Are you referring to area-based characteristics?**
The socioeconomic indicators mentioned refer to the characteristics of small areas presented in cancer atlases.
The Australian cancer atlas allows filtering using socioeconomic indicators, but this statement has been removed as it is not common in other atlases.

@@

**Reviewer suggestion:**
**Ensure all these refs are included in the list.**
References have been added to credit the original publications of the alternative displays discussed by Tobler.

### 2020-05-22 literature.Rmd
Figure numbering and section numbering, and responses to comments on doc

# Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Researchers, health authorities, governments, not-for-profits and the media are common communicators of cancer statistics. They often present statistics to the public as aggregated values for geopolitical areas. Presenting these statistics requires aggregating individual observations for the geographical units, especially for privacy protection, but also for political and policy purposes. Examples of typical geographical units include states, provinces, local government areas, and post/zip codes. Counts and incidence rates are often obtained by state health registries. For this data privacy and ethics are key concerns, even when combined to small area statistics.
This type of data is routinely collected for public health reasons and may be made available to the general public as a service to the community.

@@

# Measures mapped

@@

# Visualisation approaches

## Traditional approaches for cancer map displays {#ch:choropleth-maps}

@@

```{r choropleth-grid-create, message=FALSE, warning = FALSE, fig.cap = " A selection of choropleth cancer maps from online atlases that are publicly available. Maps of various countries were chosen: United Kingdom, Australia, Spain, USA, Canada, and display several different colour palettes and legends. These atlases are described in Table 2.", fig.align='center', fig.width=12, fig.height=10}
knitr::include_graphics("figures/choropleth_grid.png")
```

@@

## Contemporary alternatives to choropleth maps {#ch:alternatives}

### Cartograms
<!-- 1496 words inc table-->

@@

```{r ggcartograms, fig.cap = "Common alternatives to maps, showing the same information for the United States of America. The colour of each state communicates the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016: (a) contiguous cartogram distorted each state's shape by the population of the state in 2015, (b) non-contiguous cartogram preserves the shape of the cartogram, but the size now reflects the population of the state in 2015 (c) Dorling cartogram are non-contiguous, circles are used to represent each state and the population of the state determines the size in 2015 (d) hexagon tilemap (non-contiguous) uses a hexagon of equal size for each state, and colours the state by the average age-adjusted rate of incidence. ", message=FALSE, warning=FALSE}
knitr::include_graphics("figures/usa_grid.png")
```

@@

### Contiguous

@@

### Non-contiguous

@@

### Dorling

@@

### Tile Map
<!-- 173 words -->
A tile map provides a tessellated display of consistent shapes.  A similar method to a rectangular cartogram, represents each geographic area using a square. The squares are tessellated to create a grid. Each area is represented by a square of the same dimensions, each tile is usually one unit of measurement, this could be geographic regions such as states or population-based that use a consistent measure of population for each tile. Regions with over four neighbors require some necessary displacement. The tile map uses color to represent a value of a statistic for each area. A similar method to a rectangular cartogram represents each geographic area using a square of the same dimensions. There are online media sources using this method, these include [@NPR], [@FiveThirtyEight], [@WSJ], [@WP]. Tile maps may be difficult to create as they are best created manually, they require additional time and care as the number of geographic areas to include increases.

@@

### Geofacet
<!-- 166 words -->
Hafen [@IGF] introduces the term geofacet to describe a grid display of small plots. The arrangement of tiles mimics the geographic topology. Geofaceting has the functionality that a statistical plot can be constructed in each facet for each geographic area. A tile map can communicate only one value per region in a visualization, while geofaceting is a more flexible visualization for communication as it increases the amount of information displayed. Virtually any type of plot can be shown in the tile, allowing displays of multiple variables or values per geographic entity. Creating the layout of a geofacet is manual, but once created can be used for any data on that geographic base.

```{r ggtilemap, fig.cap = " Two alternative displays, tile map (left) and geofaceted map (right), showing state age-adjusted rate of incidence for lung and bronchus in the USA. In the tile map, the layout approximates spatial location, with each state being an equal box filled with color representing cancer incidence. The geo-faceted map shows bar charts laid out in a grid approximating the spatial location of the state. The maps show age-adjusted rates for males and females. This display allows the presentation of multiple variables for each geographic area.", out.width="100%"}
knitr::include_graphics("figures/gggrids.png")
```

### Multivariate displays
<!-- 222 words -->

@@

# Comparison and critique of alternative displays {#sec:compare}
<!-- 408 words -->

```{r auscartograms, fig.cap = " Cartograms showing melanoma incidence in Australia: (a) contiguous, partially population transformed, (b) non-contiguous shape preserved, (c) Dorling, (d) hexagon tile map. The contiguous cartogram has expanded the highly populated areas while preserving the full shapes of rural areas. If it accurately sized areas by population, the country would be unrecognizable. The shape-preserved is unreadable due to the small area sizes. The Dorling cartogram presents all areas but many are difficult to compare. The hexagon tile map provides a reasonable spatial distribution despite having isolated hexagons in the outback areas.", message=FALSE, warning=FALSE, cache=FALSE, fig.width=8, fig.height=6, dpi=300}
knitr::include_graphics("figures/auscartograms.png")
```

## Performance of choropleth maps or cartograms for Australia

@@

## Limitations of alternative displays

@@

## Additional considerations
<!-- 479 words -->
Cancer atlases often display supplementary graphs and plots to add more information. Additional materials such as tables, graphs, and text explanations support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers [@CPISACA]. The many displays of statistical summaries, including dot plots, bar plots, box plots, cumulative distribution plots, scatter plots, and normal probability plots, can provide alternative views of the cancer statistics. These can also display supporting statistics such as error, confidence intervals, distributions, sample or population sizes, and standard deviation.

@@

# User interaction {#ch:interacting}
<!-- 634 words -->

@@

```{r interacting, fig.cap = " Interactive controls of displays in publicly available choropleth cancer maps: (i) GUI controls for statistic, sex, age groups, continents, and cancer types for Globocan 2018 [@Globocan], (ii) Menus for variable selection and zooming on Bowel Cancer Australia Atlas, (iii) Menus for choosing variables and countries in The Cancer Atlas, (iv) Tabs for different indicators and cancer types in Global Cancer Map, (v) Menus and toggles for variable and subset selection in United States Cancer Statistics: Data Visualizations.", results = "asis", message=FALSE, warning=FALSE, cache=FALSE, fig.width=8, fig.height=6, dpi=300}
knitr::include_graphics("figures/interacting.png")
```

@@

```{r animating, fig.cap = "Two examples of advanced interactivity (and animation) in publicly available choropleth cancer maps: a. Linked maps and time-series line plots, with temporal animation in Map of Cancer Mortality Rates in Spain, b. A highly responsive magnifying glass on a map of Breast Cancer Mortality in Canada.", results = "asis", message=FALSE, warning=FALSE, cache=FALSE, fig.width=8, fig.height=6, dpi=300}
knitr::include_graphics("figures/animating.png")
```

# Conclusions
<!-- 307 words -->

>>>>

# 1. Introduction
<!-- 436 words -->
<!-- What data are we working with, why, what are the purposes -->
Researchers, health authorities, governments, not-for-profits and the media are common communicators of cancer statistics. They often present statistics to the public as aggregated values for geopolitical areas. Presenting these statistics requires aggregating individual observations for the geographical units, especially for privacy protection, but also for political and policy purposes. Examples of typical geographical units include states, provinces, local government areas, and post/zip codes. Counts and incidence rates are often obtained by state health registries. For this data privacy and ethics are key concerns, even when combined to small area statistics.
This type of data is routinely collected for public health reasons and may be made available to the general public as a service to the community.

@@

# 2. Measures mapped

@@

# 3. Visualisation approaches

## 3.1 Traditional approaches for cancer map displays {#ch:choropleth-maps}

@@

```{r choropleth-grid-create, message=FALSE, warning = FALSE, fig.cap = "Figure 1. A selection of choropleth cancer maps from online atlases that are publicly available. Maps of various countries were chosen: United Kingdom, Australia, Spain, USA, Canada, and display several different colour palettes and legends. These atlases are described in Table 2.", fig.align='center', fig.width=12, fig.height=10}
knitr::include_graphics("figures/choropleth_grid.png")
```

@@

## 3.2 Contemporary alternatives to choropleth maps {#ch:alternatives}

### 3.2.1 Cartograms
<!-- 1496 words inc table-->

@@

```{r ggcartograms, fig.cap = "Figure 2. Common alternatives to maps, showing the same information for the United States of America. The colour of each state communicates the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016: (a) contiguous cartogram distorted each state's shape by the population of the state in 2015, (b) non-contiguous cartogram preserves the shape of the cartogram, but the size now reflects the population of the state in 2015 (c) Dorling cartogram are non-contiguous, circles are used to represent each state and the population of the state determines the size in 2015 (d) hexagon tilemap (non-contiguous) uses a hexagon of equal size for each state, and colours the state by the average age-adjusted rate of incidence. ", message=FALSE, warning=FALSE}
knitr::include_graphics("figures/usa_grid.png")
```

@@

### 3.2.2 Contiguous

@@

### 3.2.3 Non-contiguous

@@

### 3.2.4 Dorling

@@

### 3.2.5 Tile Map
<!-- 173 words -->
A tile map provides a tessellated display of consistent shapes.  A similar method to a rectangular cartogram, represents each geographic area using a square. The squares are tessellated to create a grid. Each area is represented by a square of the same dimensions, each tile is usually one unit of measurement, this could be geographic regions such as states or population-based that use a consistent measure of population for each tile. Regions with over four neighbors require some necessary displacement. The tile map uses color to represent a value of a statistic for each area. A similar method to a rectangular cartogram represents each geographic area using a square of the same dimensions. There are online media sources using this method, these include [@NPR], [@FiveThirtyEight], [@WSJ], [@WP]. Tile maps may be difficult to create as they are best created manually, they require additional time and care as the number of geographic areas to include increases.

@@

### 3.2.6 Geofacet
<!-- 166 words -->
Hafen [@IGF] introduces the term geofacet to describe a grid display of small plots. The arrangement of tiles mimics the geographic topology. Geofaceting has the functionality that a statistical plot can be constructed in each facet for each geographic area. A tile map can communicate only one value per region in a visualization, while geofaceting is a more flexible visualization for communication as it increases the amount of information displayed. Virtually any type of plot can be shown in the tile, allowing displays of multiple variables or values per geographic entity. Creating the layout of a geofacet is manual, but once created can be used for any data on that geographic base.

```{r ggtilemap, fig.cap = "Figure 3. Two alternative displays, tile map (left) and geofaceted map (right), showing state age-adjusted rate of incidence for lung and bronchus in the USA. In the tile map, the layout approximates spatial location, with each state being an equal box filled with color representing cancer incidence. The geo-faceted map shows bar charts laid out in a grid approximating the spatial location of the state. The maps show age-adjusted rates for males and females. This display allows the presentation of multiple variables for each geographic area.", out.width="100%"}
knitr::include_graphics("figures/gggrids.png")
```

### 3.2.7 Multivariate displays
<!-- 222 words -->

@@

# 4. Comparison and critique of alternative displays {#sec:compare}
<!-- 408 words -->

```{r auscartograms, fig.cap = "Figure 4. Cartograms showing melanoma incidence in Australia: (a) contiguous, partially population transformed, (b) non-contiguous shape preserved, (c) Dorling, (d) hexagon tile map. The contiguous cartogram has expanded the highly populated areas while preserving the full shapes of rural areas. If it accurately sized areas by population, the country would be unrecognizable. The shape-preserved is unreadable due to the small area sizes. The Dorling cartogram presents all areas but many are difficult to compare. The hexagon tile map provides a reasonable spatial distribution despite having isolated hexagons in the outback areas.", message=FALSE, warning=FALSE, cache=FALSE, fig.width=8, fig.height=6, dpi=300}
knitr::include_graphics("figures/auscartograms.png")
```

## 4.1 Performance of choropleth maps or cartograms for Australia

@@

## 4.2 Limitations of alternative displays

@@

## 4.3 Additional considerations
<!-- 479 words -->
Cancer atlases often display supplementary graphs and plots to add more information. Additional materials such as tables, graphs, and text explanations support understanding and inference derived from maps, ensuring the message communicated will be consistent across a range of viewers [@CPISACA]. The many displays of statistical summaries, including dot plots, bar plots, box plots, cumulative distribution plots, scatter plots, and normal probability plots, can provide alternative views of the cancer statistics. These can also display supporting statistics such as error, confidence intervals, distributions, sample or population sizes, and standard deviation.

@@

# 5. User interaction {#ch:interacting}
<!-- 634 words -->

@@

```{r interacting, fig.cap = "Figure 5. Interactive controls of displays in publicly available choropleth cancer maps: (i) GUI controls for statistic, sex, age groups, continents, and cancer types for Globocan 2018 [@Globocan], (ii) Menus for variable selection and zooming on Bowel Cancer Australia Atlas, (iii) Menus for choosing variables and countries in The Cancer Atlas, (iv) Tabs for different indicators and cancer types in Global Cancer Map, (v) Menus and toggles for variable and subset selection in United States Cancer Statistics: Data Visualizations.", results = "asis", message=FALSE, warning=FALSE, cache=FALSE, fig.width=8, fig.height=6, dpi=300}
knitr::include_graphics("figures/interacting.png")
```

@@

```{r animating, fig.cap = "Figure 6. Two examples of advanced interactivity (and animation) in publicly available choropleth cancer maps: a. Linked maps and time-series line plots, with temporal animation in Map of Cancer Mortality Rates in Spain, b. A highly responsive magnifying glass on a map of Breast Cancer Mortality in Canada.", results = "asis", message=FALSE, warning=FALSE, cache=FALSE, fig.width=8, fig.height=6, dpi=300}
knitr::include_graphics("figures/animating.png")
```

# 6. Conclusions
<!-- 307 words -->
